---
title: "Analysis of of the Adverse Health and Economic Impacts of US Storms"
author: "Omar Mohamed"
date: "2019 M06 30"
output: html_document
---


## Synopsis

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage

The basic goal  is to explore the NOAA Storm Database and answer some basic questions about severe weather events. You must use the database to answer the questions below and show the code for your entire analysis. 

There is also some documentation of the database available. Here you will find how some of the variables are constructed/defined.

- National Weather Service [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)

- National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)


## libraries used

```{r,results='hide'}
library(dplyr)
library(ggplot2)

```


## Loading and Processing the Raw Data


- download Storm Data and save it in **df**


- **Property damage estimates** were entered as actual dollar amounts (the variable **PROPDMG**). But they were rounded to three significant digits, followed by an alphabetical character signifying the magnitude of the number, Alphabetical characters used to signify magnitude include “K” for thousands, “M” for millions, and “B” for billions. So I created a new variable PROPDMGEXP2 and assigned conditionally "K" = 1000, "M" = 1000000, "B" = 1000000000, in other cases 1. These variables are multiplied in the next step.

- show first 2 rows 

note : i just commented **download.file(...)** because i download it when i working on it
       and when run knit make time faster but when you work on it you must uncomment this           line and download the data
       
```{r}

data_url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
if(!file.exists("data")){dir.create("data")}
#download.file(data_url , destfile = "./data/repdata%2Fdata%2FStormData.csv.bz2")
df <- read.csv("C:/Users/omar/Documents/Exploring-NOAA-storm-database/data/repdata%2Fdata%2FStormData.csv.bz2")
df$PROPDMGEXP2 <- 1
df$PROPDMGEXP2[which(df$PROPDMGEXP == "K")] <- 1000
df$PROPDMGEXP2[which(df$PROPDMGEXP == "M" | df$PROPDMGEXP == "m")] <- 1000000
df$PROPDMGEXP2[which(df$PROPDMGEXP == "B")] <- 1000000000
head(df,n=2)

```

Show number of rows and columns

```{r}

print(dim(df))

```

show Columns Names

```{r}

colnames(df)

```

show summary of StormData

```{r}
str(df)
```

# which types of events are most harmful to population health?

**Fatalities** and **injuries** have the most impact on public health, so I will present what types of severe weather are the most dangerous.


## first working on **fatalities**

1- group the data by the **event** then summarize with the summation of **fatalities**

2- sort by descending order and select first 8 rows

```{r}

#grouped.data <- df %>% group_by(EVTYPE) %>% summarize(SumFATALITIES = sum(FATALITIES))
grouped.data <- df %>% dplyr::group_by(EVTYPE) %>% dplyr::summarize(SumFATALITIES = sum(FATALITIES))
grouped.data <- dplyr::arrange(grouped.data , desc(SumFATALITIES))
grouped.data <- grouped.data[1:8,]
```

3- plot presents FATALITIES by Event type

```{r}

ggplot(grouped.data , aes(x = EVTYPE , y = SumFATALITIES,fill=EVTYPE)) + geom_bar(stat = "identity")+coord_flip()
```


## second working on **injuries**

1- group the data by the **event** then summarize with the summation of **injuries**

2- sort by descending order and select first 8 rows

```{r}

grouped.data2 <- df %>% dplyr::group_by(EVTYPE) %>% dplyr::summarize(SumInjuries = sum(INJURIES))
grouped.data2 <- dplyr::arrange(grouped.data2 , desc(SumInjuries))
grouped.data2 <- grouped.data2[1:8,]
```

3- plot presents injuries by Event type

```{r}

ggplot(grouped.data2 , aes(x = EVTYPE , y = SumInjuries ,fill=EVTYPE)) + geom_bar(stat = "identity")+coord_flip()
```


# Which types of events have the greatest economic consequences?

1- create new variable called **SumPROPDMGEXP** and equal **(PROPDMG * PROPDMGEXP2)** 

2- group the data by the **event** then summarize with the summation of **SumPROPDMGEXP**

3- sort by descending order and select first 8 rows

```{r}

df <- dplyr::mutate(df , SumPROPDMGEXP = PROPDMG * PROPDMGEXP2)
grouped.data3 <- df %>% dplyr::group_by(EVTYPE) %>% dplyr::summarize(property_damage_estimate = sum(SumPROPDMGEXP))

grouped.data3 <- dplyr::arrange(grouped.data3 , desc(property_damage_estimate))
grouped.data3 <- grouped.data3[1:8,]
```

plot shows Property damage estimates by Event type

```{r}

ggplot(grouped.data3 , aes(x = EVTYPE , y=property_damage_estimate)) + geom_bar(stat = "identity" , aes(fill=EVTYPE))+coord_flip()

```

# summary and conclusion

**flood** has the *greatest economic consequences*. 

**Tornado** is *the most harmful to population health* because caused the most FATALITIES and injuries.
